---
title: "Threat and Demography Scores, Analysis 2"
author: "Jacob Malcom, Whitney Webber, Ya-Wei Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggthemes)
library(ICC)
library(pastecs)
library(threatdemog)
library(viridis)

# load("data/FL_scores_all.rda")
# load("data/FL_scores_multi.rda")
load("../data/FL_scores_all.rda")
load("../data/FL_scores_multi.rda")
```

## Analysis 2

#### Applying the key to a (large-ish) set of species

##### Florida species in general

First, a naive model that assumes the response variable is the threat or demography score, even though this _implies_ causality flows the opposite direction.

```{r echo=TRUE}
summary_all <- stat.desc(FL_all)
summary_all <- t(summary_all[c(1,3,4,5,8:11,13), c(-4,-1)])
summary_all <- data.frame(summary_all)
names(summary_all) <- c("n", "N/A", "min", "max", "median", "mean", "s.e.", 
                        "95% CI", "s.d.")
summary_all

```

On average, the threats have worsened and demographic status has declined for these 54 Florida species. Most of these species are not recommended for a status change, but we can compare the tallies of recommendation by score:

```{r echo = FALSE}
cat("Threat scores:\n")
tapply(FL_all$Threat,
       INDEX = FL_all$Recommendation,
       FUN = mean, na.rm = T)
table(FL_all$Threat, FL_all$Recommendation)

cat("Demography scores:\n")
tapply(FL_all$Demography,
       INDEX = FL_all$Recommendation,
       FUN = mean, na.rm = T)
table(FL_all$Demography, FL_all$Recommendation)
```

Although 'threats' and 'demography' are fundamentally different, they are related (threat changes cause demographic changes...and in some cases, demographic changes can introduce or make relevant new threats):

```{r echo = FALSE}
cat("Correlation between threat and demography scores:\n")
cor.test(FL_all$Threat, FL_all$Demography, use = "complete")
```

We can also view the threat and demographic scores as a scatterplot:

```{r echo = FALSE}
ggplot(data = FL_all, aes(x = Threat, y = Demography)) +
    geom_jitter(height = 0.2, width = 0.2, size = 4, alpha = 0.3) +
    geom_vline(xintercept = 0, color = "red", alpha = 0.5) +
    geom_hline(yintercept = 0, color = "red", alpha = 0.5) +
    theme_hc()
```

These results are discussed in greater detail in the paper.

------------

------------

##### Repeatability of scoring

We would like to know the extent to which different people read the same information and arrive at different scores. Four additional people scored a random selection of ten of the Florida species, using the same 5-yr reviews and/or Federal Register documents. First, the summaries of the threat and demography scores:

```{r echo = FALSE}
summary_mult <- stat.desc(multi)
summary_mult <- t(summary_mult[c(1,3,4,5,8:11,13),c(-2,-1)])
summary_mult <- data.frame(summary_mult)
names(summary_mult) <- c("n", "N/A", "min", "max", "median", "mean", "s.e.", 
                        "95% CI", "s.d.")
summary_mult
```

Across five scorers, these then species have negative threat scores and demographic scores are trending negative. We now want to test for observer effects. First, for threat scores:

```{r echo = TRUE}
th_m1 <- lm(Threat ~ Person + Species, data = multi)
anova(th_m1)
summary(th_m1)
qplot(resid(th_m1), 
      geom = "histogram", 
      xlab = "Residuals",
      bins = 5) + theme_hc()

ICC_thr <- ICCest(x = Species, y = Threat, data = multi)
ICC_thr
```

The threat scores are not strongly affected by observer (n.s. `person` term and small mean square), and the intra-species correlation is decent. Now for demography scores:

```{r echo = TRUE}
dem_m1 <- lm(Demography ~ Person + Species, data = multi)
anova(dem_m1)
summary(dem_m1)
qplot(resid(dem_m1), 
      geom = "histogram", 
      xlab = "Residuals",
      bins = 5) + theme_hc()

ICC_thr <- ICCest(x = Species, y = Demography, data = multi)
ICC_thr
```

The `person` term is statistically significant for demography scores, but the within-species correlations are stronger than for threat scores.

```{r echo = FALSE, fig.width = 8, fig.height = 6}
tmp <- data.frame(species = c(multi$Species, multi$Species),
                  person = c(multi$Person, multi$Person),
                  category = c(rep("Threat", length(multi$Person)),
                               rep("Demography", length(multi$Person))),
                  score = c(multi$Threat, multi$Demography))

var_plot <- ggplot(data = tmp, aes(x = species, y = score)) +
                geom_violin(fill = "gray87", colour = "gray87") +
                geom_jitter(height = 0.05,
                            width = 0.4,
                            size = 3,
                            alpha = 0.8,
                            aes(colour = person)) +
                facet_grid(. ~ category) +
                scale_color_viridis(discrete = TRUE) +
                coord_flip() +
                labs(x = "", y = "\nScore\n") +
                theme_hc() +
                theme(axis.text.y = element_text(size = 8))
var_plot
```
